Detecting the impact of land cover change on observed rainfall.

### 2019-02-28 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.Rmd
update for PeerJ paper

---
title: Detecting influences of changes in land cover on observed rainfall.
author:
  - name: Chun X. Liang
    email: summer.chunliang@gmail.com 
    affiliation: a
  - name: Floris F. van Ogtrop
    email: floris.vanogtrop@sydney.edu.au 
    affiliation: a
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: a
    footnote: Corresponding Author
address:
  - code: a
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: |
  Analysis of observational data to identify relationships between rainfall and land cover change are difficult due to multiple environmental factors that cannot be strictly controlled. In this study we present a methodology to investigate the relationship using statistical methods on data from best available sources at two sites in Australia. Gridded data of rainfall and tree cover were used as spatially corresponding local conditions. Large scale effects were represented by climatic indicators, such as SOI and IOD. Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention. 
  At a Queensland site, significant tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes. On the other hand, results from a bushfire affected NSW/VIC region suggests significant changes in the rainfall. This indicates the method works better when a abrupt change in the data can be clearly identified. The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level in both locations in data up to 2009. However, high rainfall variability and possible regrowth meant that no signifcant changes were observed in alonger time sereies to 2015.

journal: "Journal of Hydrology Regional Studies"
date: "`r Sys.Date()`"
bibliography: reflist_chap3.bib
output: 
 bookdown::pdf_book:
    base_format: rticles::elsevier_article
---
```{r packages, echo=F, warnings=F, message=F, eval=T}
require(knitr)
require(bookdown)
require(pander)
library(fields)
library(graphics)
library(maptools)
#library(plotKML)
library(raster)
library(reshape)
library(rgeos)
library(rgdal)
library(RColorBrewer)
library(tidyverse)
require(lubridate)
library(deseasonalize)
library(mgcv)
library(ncdf4)
require(gridExtra)
#library(nnet)
```

```{r setup, warning=F, message=F, echo =F}
# root dir
knitr::opts_knit$set(root.dir = 
                       "C:/Users/rver4657/ownCloud/chun/chun/manuscripts/Rainfall_Landuse_paper")
```

Introduction
==========

Land use and land cover changes can lead to changes in the local climate. Empirical and modelling studies have found cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998]. Using airborne measurement in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed. @Nair2011 reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in an observed rainfall decrease. Maximum temperatures were also found to be sensitive to land cover change in eastern Australia [@McAlpine2007].

Overall the number of empirical studies analyzing changes to rainfall due to land cover change from observational data is limited. Most of the studies mentioned previously were either model simulations, or comparisons of modelled data with observations. This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?). In addition, in many areas across the globe rainfall variability is related to complex set of interactions, of which land use change might only be a minor component.

Locally, there are two main sources that generate rainfall: moisture from advective atmospheric transport; and local evapotranspiration [@Eltahir1996; @Bosilovich2006; @Dirmeyer2009; @Gimeno2010]. The local evapotranspiration component is the component considered to be affected by land use change [@Eltahir1996]. According to @Trenberth1999, the contribution of advective moisture partially depends on the availability of external moisture and atmospheric transport. On the longer time scale, such as monthly and annually, large scale atmospheric dynamics are affected by large scale climate drivers. For example, many studies have reported significant relationships between rainfall in large parts of Australia and the El Niño-Southern Oscillation (ENSO) [@Verdon2004; @Risbey2009; @Speer2011]. In contrast, local ET is determined by local land surface characteristics, which influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both main sources. Therefore land surface is suggested to play an important role in local rainfall.

Although climate drivers demonstrate some capability to predict Australian rainfall, there is still a large amount of unexplained variance. @Westra2010 pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of annual precipitation variance. Some of the remaining variance could be due to land surface processes as suggested in studies predicting local rainfall [e.g. @Ma2011; @Zeng2012; @pitman_scale_2016; @saha_investigating_2016]. However, they are mostly based on modelling experiments and little evidence was reported from observations. However, @Pitman2004 found a good match between observations and simulated rainfall changes in southwest Western Australia ,forced by land cover change. @Timbal2006 were able to reproduce the rainfall decline in south west Australia by including land cover influence. In addition, local land use change might not be a primary, but is likely to be a secondary cause of rainfall change [@Nicholls2006].

Therefore, the aim of this study is to expand the available methodology to use empirical evidence at regional scales to investigate the cause and effect relationship between land cover change and local rainfall. More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall. To demonstrate the approach and this effect we applied the methodology to study the changes in rainfall at two locations in Queensland and NSW/Victoria where there are possible step changes in land cover change due to land clearing and bush fires. The methodology is mainly based on statistical approaches to identify changes in rainfall, which were subsequently associated with land cover change through spatial comparison.

@@

We use two areas in Australia to demonstrate the approach. In Australia, significant tree cover change has mainly occurred in the north east of the continent and on the southeast coast, as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreasing EVI post 2000. As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions. The possible EVI reduction might be due to land clearing, bush fires or drought.

```{r selreg, fig.cap="Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.", out.width="90%", echo=F}

@@

Two regions were selected where significant tree cover change was reported. The first region is located in south central Queensland to the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg). High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004]. The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg). Severe bush fires occurred in this area and the surroundings in early 2003 (see Figure \@ref(fig:figure2bushfire)). The 2003 bush fires were the largest and the worst in this area for the last 60 years [@Fire2011]. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003] and the type of species in this region. However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987]. As a result , significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.

```{r figure2bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows large area in the Kosciuszko national park has been burned. Some locations in the southwest of ACT have also experienced intensive bushfires.", out.width="90%", echo=F}
include_graphics("figures/bushfire_nswvic.jpg")
```

The two regions have different climate characteristics. The QLD region is partially grassland and partially subtropical, while the NSW/VIC region is mainly within the temperate zone, under the Köppen classification. According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double of the rainfall in the QLD region. Evapotranspiration is similar in both regions. Marine moisture and orographic effects are likely to be the main contributors to rainfall in the southeast mountain areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also different. In the Queensland region, the tree cover is sparse over most of the area. The MODIS satellite tree cover data (discussed in more detail in section 3) shows that tree cover in this region is generally below 20\% of total ground area. Grazing is the main activity in this region, with over 90\% of land used by the grazing industry [@ABARES2010]. Our starting assumption is that the main cause of the EVI  decline over large part of the region is due to land clearing. Tree cover has been cleared at a massive scale over the last decade, especially during 2002 - 2004. The Kosciuszko national park is within the NSW/VIC region. Here tree cover is denser with open or even closed forest (the tree cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant species in the alpine area are Snow Gum and large stand species such as Alpine Ash and Mountain Gum in the sub-alpine area. These trees can reach a great height but they take long time to grow. For example, Alpine Ash would need about 20 years to mature. Although land clearing is not a major issue in this region, it is vulnerable to fires and drought.

Therefore two types of land cover changes were studied. The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) [e.g. @SLATS2001; @SLATS2017] were used to investigate the time and location of land clearing in the QLD region. The MODIS burned area product, MCD45A1 [@Roy2002; @Roy2005; @Roy2008], was used to locate bush fires areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1 provides monthly burning information on all pixels, which helps to pinpoint an abrupt event. Due to the nature of the different land cover change, the post-change vegetation status in the two regions is expected to be different (see Figure \@ref(fig:figure3-tc-simple)).

@@

## the selected regions for mapping
root_dir <- "c:/users/rver4657/owncloud/chun/chun/data"
aus_reg <- readShapePoly(paste(root_dir,
                               "Geo\\aus_state2011\\STE11aAust.shp",sep="/"),
                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)

@@

fire_0301 <- raster(paste(root_dir,
          "Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.jpg",
                          sep="/"))
fire_0301nsw <- crop(fire_0301,e2)
fire_loc <- crop(reproject(fire_0301nsw,
                           CRS=CRS("+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")),
                 e222)
remove(list=c("fire_0301","fire_0301nsw"))

@@

rootdir <-"C:/users/rver4657/owncloud/chun2/data/satellite/MOD44/B"

@@

# Qld
mod_tc_r1.z <- calc(mod_tc_r11,function(x) fun(x,1))
# Qld 2011
mod_tc_r1_7y.z <- calc(mod_tc_r11,function(x) fun(x,5))
# NSW
mod_tc_r2.z <- calc(mod_tc_r22,function(x) fun(x,2))
# # NSW only 5 years after
# mod_tc_r2_5y.z <- calc(mod_tc_r22,function(x) fun(x,7))
# NSW 7 years after (2011)
mod_tc_r2_7y.z <- calc(mod_tc_r22,function(x) fun(x,5))
# # NSW 9 years after
# mod_tc_r2_9y.z <- calc(mod_tc_r22,function(x) fun(x,3))

@@

jpeg("../../figures/tc_figs.jpg", width=1280, height=960, quality=100)
par(mar=c(3,3,2,11),mfrow=c(2,2))
# Qld 7 year 2011
plot(e111, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r1_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
legend(146,-27.5,c("State","MDB"),col=c("grey","grey"),
       lwd=c(1,3),bty="n",title="Boundary",cex=1.8)
image.plot(zlim=c(-4,4), breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
           legend.only=T,horizontal=F,
           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
           legend.shrink=0.8,
           legend.width=1.5,
           legend.args=list(text=expression(paste(Delta,
                            "tree (% ground)",sep="")),
                            line=1,cex=1.3,font=2),
           axis.args=list(cex.axis=1.5),graphics.reset=F)

par(mar=c(3,9,2,5))
plot(e111, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
#            lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
#            legend.only=T,horizontal=F,
#            col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
#                  "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
#            legend.shrink=0.7,
#            legend.width=0.7,
#            legend.args=list(text=expression(paste(Delta,
#                             "tree (% ground)",sep="")),
#                             line=0.5,cex=0.9,font=2),
#            axis.args=list(cex.axis=0.8),graphics.reset=F)
# 2011 NSW
par(mar=c(3,3,1,11))
plot(e222, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r2_7y.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
           legend.only=T,horizontal=F,
           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
           legend.shrink=0.8,
           legend.width=1.5,
           legend.args=list(text=expression(paste(Delta,
                            "tree (% ground)",sep="")),
                            line=1,cex=1.3,font=2),
           axis.args=list(cex.axis=1.5),graphics.reset=F)

# FULL PERIOD NSW
par(mar=c(3,9,1,5))
plot(e222, cex.axis=1.5, font.axis=2, ylab="")
image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T)
plot(mdb_reg2,border="grey",lwd=3,add=T)
# image.plot(zlim=c(-4,4),breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
#            lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
#            legend.only=T,horizontal=F,
#            col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
#                  "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
#            legend.shrink=0.7,
#            legend.width=0.7,
#            legend.args=list(text=expression(paste(Delta,
#                             "tree (% ground)",sep="")),
#                             line=0.5,cex=0.9,font=2),
#            axis.args=list(cex.axis=0.8),graphics.reset=F)
dev.off()

jpeg("../../figures/map.jpeg", width=960, height=960)
par(mar=c(1.5,1.5,1.5,1.5))
plot(mdb_reg2,border="grey",lwd=3)
plot(aus_reg2,border="grey",add=T)
legend(150.2,-36,c("State","MDB"),col=c("grey","grey"),lwd=c(1,3),bty="n",title="Boundary",cex=1.5)
#invisible(text(getSpPPolygonsLabptSlots(aus_reg2), labels=as.character(aus_reg2$NAME), cex=0.7,adj=0.5))
box(lty=1)
plot(e111,col="red",add=T)
plot(e222,col="red",add=T)
text(147.2,-26.8,"QLD",cex=1.5)
text(148,-34.7,"NSW/VIC",cex=1.5)
dev.off()

@@

The hypothesis in this study therefore is that the effect of 2003 - 2004 land clearings in the QLD region and the 2003 bush fires in the NSW/VIC region cause a step change in the local rainfall. The actual tree cover change at the pixel level during this time was derived from the 15-year MODIS data (discussed below). The difference of tree cover before and after the land disturbance was tested using a Student's t-test. As the length of the tree cover data is shorter than available rainfall data, earlier land clearings in the QLD region cannot be identified spatially, hence they are excluded from the analysis.

## Data {#Data}

Several land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field data set (version 5). This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer then the earlier mentioned burned product MCD45A1. The data set is available on an annual time interval for the study period of 2000 - 2015. The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature [@Townshend2011]. The National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010] from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous data set. This data set, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2015. It also has a grid resolution of 250 m. The data set provides information on the final land cover types (as in 2015) and estimated trend of EVI statistics (annual mean, maximum and minimum).

Rainfall data for Australia [@Jones2009] was obtained from the Bureau of Meteorology. The data has been projected onto a national 0.05\textdegree × 0.05\textdegree grid (approximately 5 km × 5 km). This gridded data set was generated from station observations using an optimised Barnes successive correction technique. The Barnes technique combines a weighted averaging process and defined topographical information to estimate rainfall values between spatial points [@BoM2009]. The resulting data set provides additional information for data-sparse areas like central Australia but reduces information in the data-rich areas, such as southeast Australia where station density is up to 20 per 100 $\text{km}^2$. The data is available on a monthly basis from 1900 to current. Here a subset of 36 years (1979 - 2015) was used. The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible but can often found to be significant in particular months or seasons [e.g. @Otterman1990; @Gaertner2001; @Semazzi2001; @Oleson2004; @Deo2009].

>>>>

---
title: Detecting the impact of land cover change on observed rainfall.
preprint: false
author:
  - name: Chun X. Liang
    affiliation: 1
  - name: Floris F. van Ogtrop
    affiliation: 1
  - name: R. Willem Vervoort
    affiliation: 1
    corresponding: true
    email: willem.vervoort@sydney.edu.au
affiliation:
  - code: 1
    address: Sydney Institute of Agriculture, The University of Sydney, NSW 2006
abstract: >
 Analysis of observational data to identify relationships between rainfall and land cover change are difficult due to multiple environmental factors that cannot be strictly controlled. In this study we present a methodology to investigate the relationship using statistical methods on data from best available sources at two sites in Australia. Gridded data of rainfall and tree cover were used as spatially corresponding local conditions. Large scale effects were represented by climatic indicators, such as SOI and IOD. Regression analysis and step trend tests were used to assess the effect of abrupt land surface intervention. 
  At a Queensland site, significant tree cover change between 2002 - 2005 did not result in strong statistically significant precipitation changes. On the other hand, results from a bushfire affected NSW/VIC region suggests significant changes in the rainfall. This indicates the method works better when a abrupt change in the data can be clearly identified. The results from the step trend test implied a positive relationship between the tree cover and the rainfall at 0.1 significance level in both locations in data up to 2009. However, high rainfall variability and possible regrowth meant that no signifcant changes were observed in alonger time sereies to 2015. 
header-includes: >
  \usepackage{lipsum}
  \usepackage{textcomp}
bibliography: reflist_chap3.bib
date: "`r Sys.Date()`"
output: 
  rticles::peerj_article:
    base_format: bookdown::pdf_document2 # bookdown::pdf_document2 # for using \@ref()
---
```{r packages, echo=F, warnings=F, message=F, eval=T}
require(knitr)
require(bookdown)
require(pander)
library(fields)
library(graphics)
library(maptools)
library(plotKML)
library(raster)
library(reshape)
library(rgeos)
library(rgdal)
library(RColorBrewer)
library(tidyverse)
require(lubridate)
library(deseasonalize)
library(mgcv)
library(ncdf4)
require(gridExtra)
#library(nnet)
```

```{r setup, warning=F, message=F, echo =F}
# root dir
knitr::opts_knit$set(root.dir = 
                       "E:/cloudstor/chun/chun/manuscripts/PeerJRainfall_Landuse")
```

# Introduction {-}

Land use and land cover changes can lead to changes in the local climate. Empirical and modelling studies have found that cloud types and rainfall are correlated to large scale vegetation cover changes, such as deforestation in the Amazon and in the Sahel [@Chagnon2005;@Pinto2009; @Wang2009; @Mei2010; @kucharski_further_2013; @pitman_scale_2016] and afforestation in south Israel [@Otterman1990; @Ben-Gai1998]. Using airborne measurement in Western Australia, @Junkermann2009 showed a significantly higher level of aerosols over an agricultural area compared to an adjacent natural vegetation. They suggested that a modification of aerosol concentrations due to deforestation could have contributed to a reduction of local rainfall, as more, but smaller rain droplets were observed. @Nair2011 reported from the Bunny Fence Experiment in Western Australia that local land use change altered the synoptic west coast trough dynamics and surface roughness, and this resulted in an observed rainfall decrease. Maximum temperatures were also found to be sensitive to land cover change in eastern Australia [@McAlpine2007].

Overall the number of empirical studies analyzing changes to rainfall due to land cover change from observational data is limited. Most of the studies mentioned previously were either model simulations, or comparisons of modelled data with observations. This is because there are some fundamental experimental difficulties in both space (where does evaporated water reappear as rainfall?) and in time (how much time does it take for land cover change effects to appear or disappear?). In addition, in many areas across the globe, rainfall variability is related to a complex set of interactions, of which land use change might only be a minor component.

Locally, there are two main sources that generate rainfall: moisture from advective atmospheric transport; and local evapotranspiration [@Eltahir1996; @Bosilovich2006; @Dirmeyer2009; @Gimeno2010]. The local evapotranspiration component is the component considered to be affected by land use change [@Eltahir1996]. According to @Trenberth1999, the contribution of advective moisture partially depends on the availability of external moisture and atmospheric transport. On the longer time scale, such as monthly and annually, large scale atmospheric dynamics are affected by large scale climate drivers. For example, many studies have reported significant relationships between rainfall in large parts of Australia and the El Niño-Southern Oscillation (ENSO) [@Verdon2004; @Risbey2009; @Speer2011]. In contrast, local ET is determined by local land surface characteristics, which influence local scale atmospheric dynamics and hence the amount of rainfall, including contribution from both main sources.

Although climate drivers demonstrate some capability to predict Australian rainfall, there is still a large amount of unexplained variance. @Westra2010 pointed out that models based on global sea surface temperature anomalies can only predict up to 14.7\% of annual precipitation variance. Some of the remaining variance could be due to land surface processes as suggested in studies predicting local rainfall [e.g. @Ma2011; @Zeng2012; @pitman_scale_2016; @saha_investigating_2016]. However, most are based on modelling experiments and few empirical observational studies have been reported. However, @Pitman2004 found a good match between observations and simulated rainfall changes in southwest Western Australia, forced by land cover change. @Timbal2006 were able to reproduce the rainfall decline in south west Australia by including land cover influence. In addition, local land use change might not be a primary, but is likely to be a secondary cause of rainfall change [@Nicholls2006].

Therefore, the aim of this study is to use a statistical approach on rainfall data at regional scales to investigate the cause and effect relationship between land cover change and local rainfall, which is demonstrated in many modelling studies. More specifically, we hypothesize that a step change on the land cover on the surface will cause a step change in the rainfall. To demonstrate this we study changes in observed rainfall over time at a Queensland and NSW/Victoria location where there are possible step changes in land cover change due to land clearing and bush fires. The methodology uses statistical approaches to identify changes in rainfall, which are subsequently associated with land cover change through spatial comparison.

@@

In Australia, significant tree cover change has mainly occurred in the north east  and south east of the continent,  as well as in the southwest of Western Australia. According to the National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010], most of these areas have experienced decreases in the Enhanced Vegetation Index (EVI) post 2000 aws derived from satellite data. As an index for vegetation greenness, the decreasing values indicate lower biomass over time in the tree cover regions. The possible EVI reduction might be due to land clearing, bush fires or drought.

```{r selreg, fig.cap="Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest where the tree cover is denser.", out.width="90%", echo=F}

@@

Two regions were selected where significant tree cover change since 2000 was reported. The first region is located in south central Queensland partly covering the north of the Murray Darling Basin (MDB) (site 1 in Figure \@ref(fig:selreg)). High rates of land clearing have been reported in this region during the early 2000s [@SLATS2004]. The second study region is located at the border of New South Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in Figure \@ref(fig:selreg)). Severe bush fires occurred in this area  in early 2003 (see Figure \@ref(fig:figure2bushfire)). The 2003 bush fires were the largest and the worst in the last 60 years [@Fire2011]. Two thirds of Kosciuszko national park was heavily burned and regrowth was reported to be slow due to drought and cold conditions [@ABC2003], and the type of species in this region. However, in the longer term, after an early high transpiration period a recovery of pre-fire evapotranspiration would be expected [@Kuzcera1987]. For the purpose of this study, significant tree cover loss has happened in both study areas in the last decade, either permanently or temporarily.

```{r figure2bushfire, fig.cap = "Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows the large area in the Kosciuszko national park that has been burned. Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.", out.width="90%", echo=F}
include_graphics("figures/bushfire_nswvic.jpg")
```

The two regions have different characteristics. The QLD region is partially grassland and subtropical, while the NSW/VIC region is mainly within the temperate zone, under the Köppen classification. According to Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more than double the annual rainfall in the QLD region. Evapotranspiration is similar in both regions. Marine moisture and orographic effects are likely to be the main contributors to rainfall in the southeast mountain areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also different. In the Queensland region, the tree cover is sparse over most of the area. The MODIS satellite tree cover data (discussed in more detail in section 3) shows that tree cover in this region is generally below 20\% of total ground area. Grazing is the main activity in this region, with over 90\% of land used by the grazing industry [@ABARES2010]. Our starting assumption is that the main cause of the EVI  decline over large part of the region is due to land clearing. Tree cover has been cleared at a massive scale over the last decade, especially during 2002 - 2004. The reports from the Queensland Statewide Land Cover and Trees Study (SLATS) [e.g. @SLATS2001; @SLATS2017] were used to investigate the time and location of the land clearing in the QLD region.

The Kosciuszko national park is within the NSW/VIC region. Here tree cover is denser with open or even closed forest (the tree cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant species in the alpine area are Snow Gum and large stand species such as Alpine Ash and Mountain Gum in the sub-alpine area. These trees can reach a great height but they take long time to grow. For example, Alpine Ash would need about 20 years to mature. Although land clearing is not a major issue in this region, it is vulnerable to fires and drought. The MODIS burned area product, MCD45A1 [@Roy2002; @Roy2005; @Roy2008], was used to locate bush fires areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1 provides monthly burning information on all pixels, which helps to pinpoint an abrupt event.

Due to the difference in nature of the land cover change in the two regions, the post-change vegetation status is hypothesized to be different as well (see Figure \@ref(fig:figure3-tc-simple)).

@@

## the selected regions for mapping
root_dir <- "e:/cloudstor/chun/chun/data"
aus_reg <- readShapePoly(paste(root_dir,
                               "Geo\\aus_state2011\\STE11aAust.shp",sep="/"),
                         proj4string=CRS("+proj=lcc +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0"))
aus_reg2 <- gSimplify(aus_reg,0.01,topologyPreserve=T)

@@

fire_0301 <- raster(paste(root_dir,
          "Other\\bushfires\\Burnt_Area\\tmpA2003001.burndate.tif",
                          sep="/"))
fire_0301nsw <- crop(fire_0301,e2)
fire_0301nsw_t <- projectRaster(fire_0301nsw,
                           crs = "+proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=-27 +lon_0=132 +x_0=0 +y_0=0 +ellps=WGS84")
fire_loc <- crop(fire_0301nsw_t,e222)
remove(list=c("fire_0301","fire_0301nsw"))

@@

rootdir <-"E:/cloudstor/chun2/data/satellite/MOD44/B"

@@

# Qld
mod_tc_r1.z <- calc(mod_tc_r11,function(x) fun(x,1))
# Qld 2011
#mod_tc_r1_7y.z <- calc(mod_tc_r11,function(x) fun(x,5))
# NSW
mod_tc_r2.z <- calc(mod_tc_r22,function(x) fun(x,2))
# # NSW only 5 years after
# mod_tc_r2_5y.z <- calc(mod_tc_r22,function(x) fun(x,7))
# NSW 7 years after (2011)
#mod_tc_r2_7y.z <- calc(mod_tc_r22,function(x) fun(x,5))
# # NSW 9 years after
# mod_tc_r2_9y.z <- calc(mod_tc_r22,function(x) fun(x,3))

@@

jpeg("../../figures/tc_figs.jpg", width=1280, height=640, quality=100)
par(mfrow=c(1,2), mar=c(5,3,5,5))

#par(mar=c(3,9,2,5))
plot(e111, cex.axis=1.5, font.axis=2, ylab="",xlab="")
image(mod_tc_r1.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T,lwd=3,ylab="")
plot(mdb_reg2,border="grey",lwd=5,add=T,ylab="")
# FULL PERIOD NSW
par(mar=c(5,0,5,11))
plot(e222, cex.axis=1.5, font.axis=2, ylab="",xlab="")
image(mod_tc_r2.z,col=plotclr,breaks=c(-70,-20,-10,-5,0,5,10,20,70),
      legend=F,add=T)
#plot(e111,add=T)
plot(aus_reg2,border="grey",add=T, lwd=3,ylab="")
plot(mdb_reg2,border="grey",lwd=5,add=T,ylab="")
image.plot(zlim=c(-4,4), breaks=c(-4,-3,-2,-1,-0.2,0.2,1,2,3,4),
           lab.breaks=c(-70,-20,-10,-5,"no tree","0 or",5,10,20,70),
           legend.only=T,horizontal=F,
           col=c("#C51B7D","#DE77AE","#F1B6DA","#FDE0EF","white",
                 "#E6F5D0","#B8E186","#7FBC41","#4D9221"),
           legend.shrink=0.8,
           legend.width=3,
           legend.mar=5.1,
           legend.args=list(text=expression(paste(Delta,
                            "tree (% ground)",sep="")),
                            line=1,cex=1.5,font=2),
           axis.args=list(cex.axis=1.5),graphics.reset=F)
legend(147.1,-36.2,c("State","MDB"),col=c("grey","grey"),lwd=c(2,5),bty="n",title="Boundary",cex=1.5)
dev.off()

@@

The overall hypothesis is that the effect of 2003 - 2004 land clearings in the QLD region and the 2003 bush fires in the NSW/VIC region cause a step change in the local rainfall. The actual tree cover change during this time at the pixel level was derived from the 15-year MODIS data (discussed below). The difference of tree cover before and after the land disturbance was tested using a Student's t-test. As the length of the tree cover data is shorter than available rainfall data, earlier land clearing in the QLD region cannot be identified spatially, hence they are excluded from the analysis.

Data {#Data}
==============
Several land surface data sets were used in this study. The main one was the MOD44B product Global Vegetation Continuous Field data set (version 5). This data set provides estimates of percent tree cover (percentage of ground surface covered by trees) at a grid resolution of 250 m [@Townshend2011], which is finer then the earlier mentioned burned product MCD45A1. The data set is available on an annual time interval for the study period of 2000 - 2015. The tree cover data was produced from 16-day Terra MODIS Land Surface Reflectance data and Land Surface Temperature [@Townshend2011]. The National Dynamic Land Cover Dataset (DLCD) [@Lymburner2010] from the Australian Collaborative Land Use Mapping Program (ACLUMP) was used to verify the trend of vegetation cover change calculated from the previous data set. This data set, developed by Geoscience Australia and Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES), is the first nationally consistent and thematically comprehensive land cover reference for Australia. The DLCD is based on the 16-day Enhanced Vegetation Index (EVI), again from the MODIS satellite, between April 2000 and April 2015. It also has a grid resolution of 250 m. The data set provides information on the final land cover types (as in 2015) and estimated trend of EVI statistics (annual mean, maximum and minimum).

The SILO Rainfall product data for Australia was used [@SILO2001] (available online at \url{https://silo.longpaddock.qld.gov.au/})) . The data has been projected onto a national 0.05\textdegree by 0.05\textdegree grid (approximately 5 km by 5 km). This gridded data set was generated from station observations using spline interpolation and kriging [@SILO2001]. The data has been compared to other gridded products and observed data and is generally of high quality [@Beesley2009; @Tozer2012]. The data is available on a daily and monthly basis from 1889 to current. Here a subset of 36 years (1979 - 2015) was used. The study was conducted on monthly data, as a land cover change effect on annual rainfall might be negligible, but can often  be significant in particular months or seasons [e.g. @Otterman1990; @Gaertner2001; @Semazzi2001; @Oleson2004; @Deo2009].
